---
title: Polygenic Risk Score (PRS) Introduction 201 
subtitle: basic PRS calculation and performance evaluation
author: Drs. Lei Sun, Wei Deng, Yanyan Zhao
institute:
  - Department of Statistical Sciences, FAS  
  - Division of Biostatistics, DLSPH  
  - University of Toronto    
date: "`r format(Sys.time(), '%d %B, %Y')`"

output: 
  beamer_presentation: 
    pandoc_args: ["--extract-media", "./extracted-image"]
# output: word_document
# output: html_document
# output: pdf_document

header-includes:
 - \usepackage{fancyhdr}
 - \usepackage{amsmath,latexsym}
 - \hypersetup{colorlinks=true, linkcolor=blue}
---

```{r setup, include=FALSE}
# knitr::opts_chunk$set(echo = TRUE)
# knitr::opts_chunk$set(out.width = '70%') 
set.seed(1234) 
library(knitr)
library(kableExtra)
library(plotly)
library(fields) 
```

## At the end of this lecture, a **deeper** understanding of 
\begin{itemize}  
\item the complexity of constructing a good PRS even under the simplest setting {\it without} LD or any heterogeneities..
\item the trouble introduced by false positives, due to multiple hypothesis testing and low power.
\item `the more is not always better' statement: PRS based on 6 gw-significant SNPs vs. 66 0.01-significant SNPS.
\item the various over-fitting or selection biases, and winner’s curse in $\hat \beta$ for both false positives and true positives.
\end{itemize}

## \footnotesize Recall the illustrative `polygenic' model simulation study 
\vspace{6pt}

10 out 5000 indep.\ SNPs with **varying `moderate-large' effects** are truly associated with $Y$ (**all $\beta=0.3$ but MAF vary**).  
\small
$$Y_i=\sum_{j=1}^{10} \beta_jG_{ij} + e, \text{ where } \beta_j=0.3$$ 
$$\text{ MAF} \sim \text{  Unif(0.05,0.5)}, \: e\sim N(0,1).$$  
\vspace{6pt}

\scriptsize
```{r,echo=TRUE,eval=FALSE}
nsnp.true=10 # number of truly associated SNPs
beta.true=0.3 # "large" effect (also MAF, the error term, and the sample size)
```
\vspace{6pt}

```{r,echo=FALSE,out.width = '50%',fig.align = "center"}
set.seed(101) 
nsample=1000;nsnp=5000 
G=matrix(-9,nrow = nsample,ncol = nsnp)
maf=runif(nsnp,min=0.05,max=0.5)
maf.hat=rep(-9,nsnp)
nsnp.true=10; beta.true=0.3 # +-sign later
beta=c(rep(beta.true,nsnp.true),rep(0,(nsnp-nsnp.true)))
betaG=rep(0,nsample)
for(j in 1:nsnp){ 
  nG=rmultinom(1,size=nsample,prob=c((1-maf[j])^2,2*maf[j]*(1-maf[j]), maf[j]^2))
  maf.hat[j]=(2*nG[3]+nG[2])/(2*nsample)
  G[,j]=sample(c(rep(0,nG[1]),rep(1,nG[2]),rep(2,nG[3]))) # shuffle the G; no LD
  betaG=betaG+beta[j]*G[,j]
}
beta.0=0;sigma=1;e=rnorm(nsample,mean=0,sd=sigma) 
Y=beta.0+betaG+e

sumstat=matrix(-9,nrow=nsnp,ncol=7)
colnames(sumstat)=c("MAF", "MAF.hat", "beta", "beta.hat", "se", "Z.value", "p.value")
for(j in 1:nsnp){
  fit=lm(Y~G[,j])
  sumstat[j,]=c(maf[j],maf.hat[j],beta[j], summary(fit)$coefficients[2,])
}
hist(Y,freq=F)
curve(dnorm(x,mean=mean(Y), sd=sqrt(var(Y))), add=T,lwd=2,col="blue")
```

## Recall (NOT realistic!) $PRS_{i,oracle}=\sum_{j=1}^{J=10} 0.3 \cdot G_{ij}$ 
\footnotesize

**The MAF of the 10 truly associated SNPs**  
```{r,echo=FALSE}
round(maf[1:nsnp.true],3)
```

**The SNP heritability vary, despite all $\beta_j=0.3$**  
```{r,echo=FALSE}
V.G.loci=beta[1:nsnp.true]^2*(2*maf[1:nsnp.true]*(1-maf[1:nsnp.true]))
V.G=sum(beta[1:nsnp.true]^2*maf[1:nsnp.true])
V.e=sigma^2 
h2.loci=V.G.loci/(V.G+V.e)
round(h2.loci,3) # h2 contributed by each individual SNP
```

**The true heritability, $h^2$**  
```{r,echo=FALSE}
V.G=sum(beta[1:nsnp.true]^2*(2*maf[1:nsnp.true]*(1-maf[1:nsnp.true])))
V.e=sigma^2
h2=V.G/(V.G+V.e)
round(h2,3)
```

```{r,echo=FALSE,out.width = '60%',fig.align = "center"}
# not realistic: if we knew the true value of beta_j and which which j to use, why GWAS!
PRS.oracle=rep(0,nsample)  
for (i in 1:nsample) 
  for (j in 1:nsnp.true) 
    PRS.oracle[i] = PRS.oracle[i]+beta[j]*G[i,j] 
par(mfrow=c(1,2))
hist(PRS.oracle) # not quite normal as J=10 here
plot(PRS.oracle,Y) # much more predictive than individual SNPs

Y.STD=(Y-mean(Y))/sqrt(var(Y))
PRS.oracle.STD=(PRS.oracle-mean(PRS.oracle))/sqrt(var(PRS.oracle))
```

## \footnotesize Recall the (highly significant) association between PRS.orcale and the trait
\tiny
```{r,echo=FALSE,out.width = '60%',fig.align = "center"}
summary(lm(Y.STD~PRS.oracle.STD))
```

## Recall the liability model, and the case-control stratified PRS distributions   
\tiny
```{r,echo=FALSE,out.width = '90%',fig.align = "center"}
case.index=which(Y.STD>1);control.index=which(Y.STD<=1) # 1 is a subjective choice
Y.cc=rep("0-Control",nsample); Y.cc[case.index]="1-Case"
color.index=rep("black",nsample);color.index[case.index]="red" 
par(mfrow=c(1,2))
plot(PRS.oracle.STD,Y.STD,col=color.index)
abline(h=1,col="grey"); abline(v=1,col="blue")

hist(PRS.oracle.STD[control.index],col=rgb(0,0,0,0.1),freq=F,breaks=7,xlim=c(-4,4),main="",xlab="PRS.oracle.STD")
curve(dnorm(x,mean=mean(PRS.oracle.STD[control.index]), sd=sqrt(var(PRS.oracle.STD[control.index]))), add=T,lwd=2,col="black")
hist(PRS.oracle.STD[case.index],col=rgb(1,0,0,0.1),breaks=7,freq=F,add=T)
curve(dnorm(x,mean=mean(PRS.oracle.STD[case.index]), sd=sqrt(var(PRS.oracle.STD[case.index]))),add=T,lwd=2,col="red")
```

## Recall the ROC curve and AUC using $PRS_{oracle}$
\footnotesize
```{r,echo=FALSE,out.width = '80%',fig.align = "center"}
ncase=sum(Y.cc=="1-Case") # total number of cases
ncontrol=sum(Y.cc=="0-Control") # total number of controls
a=seq(4,-2.5,by=-0.5)
P=rep(-9,length(a));TP=rep(-9,length(a));FP=rep(-9,length(a))
sensitivity=rep(-9,length(a)); specificity.1=rep(-9,length(a))
for (k in 1:length(a)) {
 PRS.threshold=a[k]
 P[k]=sum(PRS.oracle.STD>PRS.threshold) # number of Positives at this threshold
 TP[k]=sum(Y.cc[PRS.oracle.STD>PRS.threshold]=="1-Case") # True Positives
 FP[k]=sum(Y.cc[PRS.oracle.STD>PRS.threshold]=="0-Control") # False Positives
 sensitivity[k]=TP[k]/ncase # sensitivity, estimated by TP/ncase
 specificity.1[k]=FP[k]/ncontrol # 1-specificity, estimated by 1 - true negatives/ncontrol
}
plot(specificity.1,sensitivity,type="b",pch=20,col="blue",xlab="1 - Specificity = False Positives/ncontrols",ylab="Sensitivity = True Positives/ncases")
abline(0,1)
auc=0  
for (k in 2:length(a)) # Trapezoid's method height*(base1+base2)/2
  auc = auc + (specificity.1[k]-specificity.1[k-1])*(sensitivity[k]+sensitivity[k-1])/2
c("AUC of ROC.oracle=",round(auc,3))
```

## Recall the \color{red}{\bf BUT},

$$ {\bf PRS_{i, oracle}}=\sum_{j=1}^{J=10} \beta_j(=0.3) G_{ij}\:\: \color{red}{\bf \text{\bf is NOT } PRS_{i,parctice}!}$$  

\begin{itemize}
\item $J$ is unknown, to be determined  
\item $\beta_j$ is unknown, to be estimated  
\item  $G_{ij}$ {\bf cannot be directly from the same data used to infer $J$ and $\beta_j$}.  
\item[] Otherwise: over-fitting/double-dipping/data-dredging/p-hacking/selection-bias!  
\item Not to mention LD and other considerations in real data settings.
\end{itemize}
\vspace{12pt}  

## What we have: $\hat \beta$, $Z$ and p-values
```{r,echo=FALSE,out.width = '90%',fig.align = "center"}
par(mfrow=c(2,2))
hist(sumstat[,"beta.hat"],freq=F)
curve(dnorm(x,mean=0, sd=(sqrt(var(sumstat[,"beta.hat"])))), add=T,lwd=2,col="blue")
hist(sumstat[,"Z.value"],freq=F)
curve(dnorm(x,mean=0, sd=1), add=T,lwd=2,col="blue")
hist(sumstat[,"p.value"],freq=F)
abline(h=1,col="blue")
plot(-log10((seq(1,nsnp,1)-0.5)/nsnp), -log10(sort(sumstat[,"p.value"])),ylab="-log(p-value) Obs", xlab="-log(p-value) Exp",main="QQ-plot of p-value")
abline(0,1,lwd=2,col="blue")
```

## \footnotesize Recall the sumstat (the true beta and MAF are here thanks to simulation)
\tiny
```{r,echo=FALSE}
sumstat[1:30,]
```

## Determine $J$ and Estimate $\beta_j$ using GW significance level 
\scriptsize
```{r,echo=TRUE,out.width = '45%',fig.align = "center"}
J.index=which(sumstat[,"p.value"]<=0.05/nsnp) #10^-5 here for 5000 SNPs
J.index # the index for the significant SNPs
c(length(J.index),sum(J.index<=nsnp.true)) # positives, true positives
round(sumstat[J.index,"beta.hat"],2) # beta estimates for the significant SNPs
hist(sumstat[J.index,"beta.hat"]); abline(v=beta.true,col="blue")
```

## A less stringent significance level, say $\alpha=0.01$?  

\small
\centerline{\bf Trade-off: between false positives (56) and power (10 out 10)}  
\vspace{3pt} 
\tiny
```{r,echo=TRUE,out.width = '50%',fig.align = "center"}
J.index=which(sumstat[,"p.value"]<=0.01)
J.index # the index for significant SNPs
c(length(J.index),sum(J.index<=nsnp.true)) # positives, true positives
```
\vspace{6pt}

\small
\centerline{\bf Trouble ahead: {\color{red}$|\hat \beta_j$| of the FP SNPs} are competitive!}

\tiny
```{r,echo=FALSE,out.width = '50%',fig.align = "center"}
hist(abs(sumstat[J.index[-(1:10)],"beta.hat"]),col=rgb(1,0,0,0.1),freq=T,xlim=c(0,0.5),main="Red=FP SNPs vs. Black=TP SNPs",xlab="abs(Estimates of beta)")
hist(abs(sumstat[J.index[1:10],"beta.hat"]),col=rgb(0,0,0,0.1),freq=T,add=T)
```

## Top associated SNPs for $PRS_i=\sum_{j=1}^{J} \hat \beta_j G_{ij}$: \color{red}$J$  
**Genome-wide significance level**  ($\alpha=10^{-5}$ here for 5000 SNPs)  
    
- $J=6$  
- find only 6 out 10 truly associated SNPs 
- but 0 false positives  
\vspace{6pt}  
  
**A less stringent significance level** ($\alpha=0.01$)  
  
- $J=66$
- find all 10 truly associated SNP
- but 56 false positives 
\vspace{6pt}  

\small
\color{red}Live Quiz 1: Which $\alpha$ threshold will leads to a better PRS (higher AUC)?  
  
A: using 6 SNPS with GW significance  
B: using 66 SNPs with p< 0.01  
C: ~same

## Effect size estimates in $PRS_i=\sum_{j=1}^{J} \hat \beta_j G_{ij}$: \color{red}$\hat \beta_j$  

**Genome-wide significance level** 
\tiny
```{r,echo=TRUE,out.width = '60%',fig.align = "center"}
J.index=which(sumstat[,"p.value"]<=0.05/nsnp) 
round(sumstat[J.index,"beta.hat"],2)
```
\vspace{6pt}  
  
\normalsize
**A less stringent significance level**  
\tiny
```{r,echo=TRUE,out.width = '60%',fig.align = "center"}
J.index=which(sumstat[,"p.value"]<=0.01) 
round(sumstat[J.index,"beta.hat"],2)
```
\vspace{6pt}    

\normalsize  
That was too easy! **More considerations** later:  

- Winner's curse (a result of low power) and the MAF connection  
- heterogeneity and transportability  
- LD   

## \footnotesize Quiz: patterns for the signicant SNPs?  

\small
**MAF of the 10 truly associated SNPs**
```{r,echo=FALSE,out.width = '60%',fig.align = "center"}
round(sumstat[1:nsnp.true,"MAF"],2)
```
\vspace{12pt}

**MAF of the 6 significant SNPs at the GW level**
```{r,echo=FALSE,out.width = '60%',fig.align = "center"}
round(sumstat[which(sumstat[,"p.value"]<=0.05/nsnp),"MAF"],2)
```
\vspace{12pt}

**Sample estimates of the 6 significant SNPs at the GW level**
```{r,echo=FALSE,out.width = '60%',fig.align = "center"}
round(sumstat[which(sumstat[,"p.value"]<=0.05/nsnp),"MAF.hat"],2)
```

## \footnotesize Quiz cont'd, -log(GWAS p-value) vs. $\hat \beta_j$   
\vspace{12pt}

\small \centerline{(Red = FP SNPs vs. Black = TP SNPs)} 
\tiny
```{r,echo=FALSE,out.width = '70%',fig.align = "center"}
par(mfrow=(c(1,2)))
J.index=which(sumstat[,"p.value"]<=0.01) 
temp=rep("red",length(J.index));temp[1:10]="black" 
plot(sumstat[J.index,"beta.hat"],-log10(sumstat[J.index,"p.value"]),main="The 66 significant SNPs at the 0.01 level",col=temp,ylim=c(0,10))
temp=rep("red",nsnp);temp[1:10]="black" 
plot(sumstat[,"beta.hat"],-log10(sumstat[,"p.value"]),main="All SNPs at the 0.01 level",col=temp)
```
\vspace{1cm}

\scriptsize
<!-- Sun and Bull (2005). *Genetic Epidemiology*. [Reduction of selection bias in genome-wide genetic studies by resampling.](https://pubmed.ncbi.nlm.nih.gov/15761913/)   -->
Sun et al. (2011). *Human Genetics*. [BR-squared: a practical solution to the winner's curse in genome-wide scans.](https://pubmed.ncbi.nlm.nih.gov/21246217/)

## Can we construct $PRS_i=\sum_{j=1}^{J} \hat \beta_j G_{ij}$ now?  
What we have using GW $\alpha=10^{-5}$ (for 5000 SNPs):  
\vspace{12pt}

**The number of SNPs, J**  
```{r,echo=FALSE}
J.index=which(sumstat[,"p.value"]<=0.05/nsnp)
length(J.index) # J
```
\vspace{12pt}

**Which specific SNPs**  
```{r,echo=FALSE}
J.index
```
\vspace{12pt}

**$\hat \beta_j$ of these SNPS**  
```{r,echo=FALSE}
round(sumstat[J.index,"beta.hat"],2) # beta estimates for the J SNPs
```

## WRONG if using $G_{ij}$ from the same data!
$$PRS_{i,wrong}=\sum_{j=1}^{6} \hat \beta_j G_{ij}$$
\vspace{6pt}
\tiny
```{r,echo=FALSE,out.width = '60%',fig.align = "center"}
PRS.wrong=rep(0,nsample) 
for (i in 1:nsample) {
  for (j in 1:length(J.index))
    PRS.wrong[i] = PRS.wrong[i]+sumstat[J.index[j],"beta.hat"]*G[i,J.index[j]]
}
PRS.wrong.STD=(PRS.wrong-mean(PRS.wrong))/sqrt(var(PRS.wrong))
par(mfrow=c(1,2))
hist(PRS.wrong.STD)
plot(PRS.wrong.STD,Y.STD,col=color.index);abline(a=lm(Y.STD~PRS.wrong.STD)$coef[1],b=lm(Y.STD~PRS.wrong.STD)$coef[2],lty=2)  
```
\vspace{12pt}

\normalsize
This PRS.wrong appears to be ~normally distributed and highly predictive of the outcome, BUT **due to over-fitting!**

## If you are not fully convinced of \color{red}{the over-fitting issue:}

\small
Using $\alpha=0.01$ BUT exclude the all the true positives. That is, **using only the following 56 false positive SNPs to construct PRS:** 
\tiny
```{r,echo=FALSE}
J.index=which(sumstat[,"p.value"]<=0.01)
J.index=J.index[11:length(J.index)]
length(J.index) # J
J.index
```
\vspace{8pt}

\small
**Their effect size sample estimates:**
\tiny
```{r,echo=FALSE}
round(sumstat[J.index,"beta.hat"],2)
```
\vspace{8pt}

\small
**Their MAF sample estimates:**   
\tiny
```{r,echo=FALSE}
round(sumstat[J.index,"MAF.hat"],2)
```

## PRS, using only null SNPs, is predictive: clearly WRONG! 
\vspace{12pt}

\tiny
```{r,echo=FALSE,out.width = '80%',fig.align = "center"}
PRS.wrong=rep(0,nsample) 
for (i in 1:nsample) {
  for (j in 1:length(J.index))
    PRS.wrong[i] = PRS.wrong[i]+sumstat[J.index[j],"beta.hat"]*G[i,J.index[j]]
}
PRS.wrong.STD=(PRS.wrong-mean(PRS.wrong))/sqrt(var(PRS.wrong))
par(mfrow=c(1,2))
hist(PRS.wrong.STD, main="PRS from using 56 false positive SNPs")
plot(PRS.wrong.STD,Y.STD,col=color.index);abline(a=lm(Y.STD~PRS.wrong.STD)$coef[1],b=lm(Y.STD~PRS.wrong.STD)$coef[2],lty=2)  
```
\vspace{12pt}

## Obtaining a significant result $\neq$ a correct result! 
\small
\bf This PRS.wrong, constructed from the 56 null SNPs, is actually more significantly associated with the phenotype than PRS.oracle: {\color{red}{clearly wrong!}}  
\vspace{12pt}

\tiny
```{r,echo=FALSE,out.width = '60%',fig.align = "center"}
summary(lm(Y.STD~PRS.wrong.STD))
```

## Cannot be overemphasized  

A predictive (and normally distributed) PRS, on its own, is not evidence for correct PRS calculation!  
\vspace{24pt}

**Remember the superscripts** in your PRS calculation: 
$$\hat \beta_j^{\color{red}{external\: (base,\: discovery)}} \times G_{ij}^{\color{red}{my.data\: (target,\: validation)}}$$  
\vspace{24pt}

Surely we will not make this rookie mistake! BUT,  
  
**over-fitting can appear in other (subtler) forms**,  e.g.  
overlapping samples between the external and my data, or  
pleitropy studies of multiple phenotypes from a single sample 


## How to construct \color{red}$PRS_i=\sum_{j=1}^{J} \hat \beta_j G_{ij}$ then? e.g. **The simplest scenario**    
  
Obtain the $J$ and $\hat \beta_j$ from **an external data set**.  
\vspace{24pt} 

The external data set resembles our own data set perfectly, i.e.  
**no heterogeneity** in population, sampling design etc.   
\vspace{24pt}  

Calculate the $PRS_i$ for each individual $i$ in our own sample for prediction:   
$$PRS_i^{my.data}=\sum_{j=1}^{J} \hat \beta_j^{external} G_{ij}^{my.data}$$
  
## Simulate an independent set of data, my.data  

\scriptsize
```{r,echo=TRUE,out.width = '60%',fig.align = "center"}
# Assume the previous data was the external data
# the external model was 
# nsnp=5000; nsnp.true=10; beta.true=0.3; beta.0=0; sigma=1

# Use the SAME MODEL but a DIFFERNT SEE to generate new independent data

set.seed(102)

my.nsample=1000 # my. is for my own data for prediction or validation
my.nsnp=nsnp # no heterogeneity: the same number of SNPs
my.maf=maf # no heterogeneity: the same MAF as before 
my.nsnp.true=nsnp.true # no heterogeneity: the same number of truly associated SNPs as before
my.beta.true=beta.true # no heterogeneity: the same effect size as before
my.beta=c(rep(my.beta.true,my.nsnp.true),rep(0,(my.nsnp-my.nsnp.true))) 
```

```{r,echo=FALSE,out.width = '60%',fig.align = "center"}
my.maf.hat=rep(-9,my.nsnp) 
my.G=matrix(-9,nrow = my.nsample,ncol = nsnp)
my.betaG=rep(0,my.nsample)

for(j in 1:nsnp){ 
  my.nG=rmultinom(1,size=my.nsample,prob=c((1-my.maf[j])^2,2*my.maf[j]*(1-my.maf[j]), my.maf[j]^2))
  my.maf.hat[j]=(2*my.nG[3]+my.nG[2])/(2*my.nsample) # MAF estimated from the sample 
  my.G[,j]=sample(c(rep(0,my.nG[1]),rep(1,my.nG[2]),rep(2,my.nG[3]))) 
  my.betaG=my.betaG+my.beta[j]*my.G[,j]
}
my.beta.0=0;my.sigma=1;my.e=rnorm(my.nsample,mean=0,sd=my.sigma)
my.Y=my.beta.0+my.betaG+my.e # the phenotype vector

my.sumstat=matrix(-9,nrow=nsnp,ncol=7)
colnames(my.sumstat)=c("MAF", "MAF.hat", "beta", "beta.hat", "se", "Z.value", "p.value")
for(j in 1:nsnp){
  my.fit=lm(my.Y~my.G[,j])
  my.sumstat[j,]=c(my.maf[j],my.maf.hat[j],my.beta[j],summary(my.fit)$coefficients[2,])
}
```

## 

**Using the same model as above (no heterogeneity)**:  
10 out 5000 SNPs are truly associated with `moderate-large' effect  
$$Y_i^{my.data}=\sum_{j=1}^{10} 0.3\times G_{ij}^{my.data} + e^{my.data}, \text{ where } e^{my.data}\sim N(0,1) $$ 
\vspace{6pt}

**MAFs stay the same as the external data (no heterogeneity)**,  
and recall the MAFs of the 10 truly associated SNPs:
```{r,echo=FALSE}
round(my.maf[1:nsnp.true],2)
```
\vspace{12pt}

The sample size does not have to be the same, but for now we use my.nsample=1000.

## \footnotesize EDA (exploratory data analysis) of my.data  
```{r,echo=FALSE, out.width = '90%',fig.align = "center"}
par(mfrow=c(2,2))
hist(my.Y,freq=F)
j=1
plot(my.G[,j],my.Y,main=paste("SNP 1, MAF=",round(my.sumstat[j,"MAF"],2), ", MAF.hat=",round(my.sumstat[j,"MAF.hat"],2), sep=""),xlab="my.G SNP 1")
my.fit=lm(my.Y~my.G[,j]); abline(a=my.fit$coef[1],b=my.fit$coef[2],lty=2)
j=2
plot(my.G[,j],my.Y,main=paste("SNP 2, MAF=",round(my.sumstat[j,"MAF"],2), ", MAF.hat=",round(my.sumstat[j,"MAF.hat"],2), sep=""),xlab="my.G SNP 2")
my.fit=lm(my.Y~my.G[,j]); abline(a=my.fit$coef[1],b=my.fit$coef[2],lty=2)
j=3
plot(my.G[,j],my.Y,main=paste("SNP 3, MAF=",round(my.sumstat[j,"MAF"],2), ", MAF.hat=",round(my.sumstat[j,"MAF.hat"],2), sep=""),xlab="my.G SNP 3")
my.fit=lm(my.Y~my.G[,j]); abline(a=my.fit$coef[1],b=my.fit$coef[2],lty=2)
```

##  
\footnotesize 
**my.sumstat** (the true beta and MAF are here thanks to simulation)
\tiny
```{r,echo=FALSE}
my.sumstat[1:13,]
```

\footnotesize 
compared with the **ex.sumstat** from the external data
\tiny
```{r,echo=FALSE}
sumstat[1:13,]
```

## Finally, \color{red}$PRS_i=\sum_{j=1}^{J} \hat \beta_j G_{ij}$  
\footnotesize
**Using GW threshold on the external data**  
$$my.PRS_{GW}=\sum_{j=1}^{6\: Positives \: (all\:  TP)} \hat \beta_j^{external}\times G_{ij}^{my.data}$$ 
**Using the $\alpha=0.01$ threshold on the external data**  
$$my.PRS_{.01}=\sum_{j=\{1:10,324,...,4935\}}^{66\: Positives \: 
(10\: TP)} \hat \beta_j^{external}\times G_{ij}^{my.data}$$
\vspace{3pt}

**Using the $\alpha=0.01$ threshold on the external data AND use only the false positives** (made possible by the simulation and should NOT be predictive when calculated correctly!)
$$my.PRS_{.01.null}=\sum_{j=\{324,...,4935\}}^{56\: False\: Positives} \hat \beta_j^{external}\times G_{ij}^{my.data}$$
**The oracle one** (made possible by the simulation)
$$my.PRS_{Oracle}=\sum_{j=1}^{all\: 10\: causal\: ones} 0.3 \times G_{ij}^{my.data}$$

\tiny
```{r,echo=FALSE,out.width = '50%',fig.align = "center"}
my.PRS.GW=rep(0,my.nsample) 
my.PRS.01=rep(0,my.nsample)
my.PRS.01.null=rep(0,my.nsample)
my.PRS.oracle=rep(0,my.nsample) 
# J index & beta.hat from the external data 
for (i in 1:my.nsample) {
  J.index=which(sumstat[,"p.value"]<=0.05/nsnp) 
  for (j in 1:length(J.index)) 
    my.PRS.GW[i] = my.PRS.GW[i]+sumstat[J.index[j],"beta.hat"]*my.G[i,J.index[j]]
  J.index=which(sumstat[,"p.value"]<=0.01) 
  for (j in 1:length(J.index))
    my.PRS.01[i] = my.PRS.01[i]+sumstat[J.index[j],"beta.hat"]*my.G[i,J.index[j]]
  J.index=J.index[11:length(J.index)]
  for (j in 1:length(J.index))
    my.PRS.01.null[i] = my.PRS.01.null[i]+sumstat[J.index[j],"beta.hat"]*my.G[i,J.index[j]]
  for (j in 1:my.nsnp.true)  
    my.PRS.oracle[i] = my.PRS.oracle[i]+my.beta[j]*my.G[i,j] 
}
```

## \footnotesize Raw and standardized (STD) of the PRSs constructed
\tiny  
```{r,echo=FALSE,out.width = '90%',fig.align = "center"}
my.Y.STD=(my.Y-mean(my.Y))/sqrt(var(my.Y)) # sample-specific rescaling can be a problem 
my.case.index=which(my.Y.STD>1);my.control.index=which(my.Y.STD<=1) # in the presence of heterogeneity between two datasets 
my.color.index=rep("black",my.nsample);my.color.index[my.case.index]="red" 

my.PRS.GW.STD=(my.PRS.GW-mean(my.PRS.GW))/sqrt(var(my.PRS.GW)) # this sample rescaling is not possible without a sample of my.data, using the external data can be tricky as well: heterogeneity and PRS_external overting
my.PRS.01.STD=(my.PRS.01-mean(my.PRS.01))/sqrt(var(my.PRS.01))
my.PRS.01.null.STD=(my.PRS.01.null-mean(my.PRS.01.null))/sqrt(var(my.PRS.01.null))
my.PRS.oracle.STD=(my.PRS.oracle-mean(my.PRS.oracle))/sqrt(var(my.PRS.oracle))

par(mfrow=c(2,4))
hist(my.PRS.oracle);hist(my.PRS.GW);hist(my.PRS.01);hist(my.PRS.01.null)
hist(my.PRS.oracle.STD);hist(my.PRS.GW.STD);hist(my.PRS.01.STD);hist(my.PRS.01.null.STD)
```

## Performance of the PRSs, from the association perspective
\tiny  
```{r,echo=FALSE,out.width = '90%',fig.align = "center"}
par(mfrow=c(1,4))
plot(my.PRS.oracle.STD,my.Y.STD,col=my.color.index)
fit=lm(my.Y.STD~my.PRS.oracle.STD);abline(a=fit$coef[1],b=fit$coef[2],lty=2)
temp=summary(fit)$coefficients[2,]
plot(my.PRS.GW.STD,my.Y.STD,col=my.color.index)
fit=lm(my.Y.STD~my.PRS.GW.STD);abline(a=fit$coef[1],b=fit$coef[2],lty=2)
temp=rbind(temp,summary(fit)$coefficients[2,])
plot(my.PRS.01.STD,my.Y.STD,col=my.color.index)
fit=lm(my.Y.STD~my.PRS.01.STD);abline(a=fit$coef[1],b=fit$coef[2],lty=2)
temp=rbind(temp,summary(fit)$coefficients[2,])
plot(my.PRS.01.null.STD,my.Y.STD,col=my.color.index)
fit=lm(my.Y.STD~my.PRS.01.null.STD);abline(a=fit$coef[1],b=fit$coef[2],lty=2)
temp=rbind(temp,summary(fit)$coefficients[2,])
c("slope.hat", round(temp[1,1],3), round(temp[2,1],3), round(temp[3,1],3), round(temp[4,1],3))
c("Z.value", round(temp[1,3],3), round(temp[2,3],3), round(temp[3,3],3), round(temp[4,3],3))
c("p.value", signif(temp[1,4],digits=4),signif(temp[2,4],3), signif(temp[3,4],3), round(temp[4,4],3))
```

##
**$PRS{.01.null}$ is NOT associated with the trait as expected!**  
\tiny
```{r,echo=FALSE,out.width = '80%',fig.align = "center"}
summary(lm(my.Y.STD~my.PRS.01.null.STD))$coefficients
```
\vspace{6pt}  
\normalsize
**$PRS_{oracle}$ is the best, but $PRS_{oracle}$ is not realistic!**  
\tiny
```{r,echo=FALSE,out.width = '80%',fig.align = "center"}
summary(lm(my.Y.STD~my.PRS.oracle.STD))$coefficients
```
\vspace{6pt}   
\normalsize
**$PRS_{GW}$ is significantly associated with the phenotype, but less so than $PRS_{oracle}$ as it should be**
\tiny  
```{r,echo=FALSE,out.width = '80%',fig.align = "center"}
summary(lm(my.Y.STD~my.PRS.GW.STD))$coefficients
```
\vspace{6pt}   
\normalsize
**More is not necessarily better: $PRS_{.01} (J=66)$ is worse than $PRS_{GW} (J=6)$ \color{red}{in this case}** 
\tiny  
```{r,echo=FALSE,out.width = '80%',fig.align = "center"}
summary(lm(my.Y.STD~my.PRS.01.STD))$coefficients
```

## Case-control stratified distributions of the different PRSs
\tiny
```{r,echo=FALSE,out.width = '90%',fig.align = "center"}
par(mfrow=c(2,2))

hist(my.PRS.oracle.STD[my.control.index],col=rgb(0,0,0,0.1),freq=F,breaks=7,xlim=c(-4,4),ylim=c(0,0.5), main="my.PRS.oracle.STD",xlab="")
curve(dnorm(x,mean=mean(my.PRS.oracle.STD[my.control.index]), sd=sqrt(var(my.PRS.oracle.STD[my.control.index]))), add=T,lwd=2,col="black")
hist(my.PRS.oracle.STD[my.case.index],col=rgb(1,0,0,0.1),breaks=7,freq=F,add=T)
curve(dnorm(x,mean=mean(my.PRS.oracle.STD[my.case.index]), sd=sqrt(var(my.PRS.oracle.STD[my.case.index]))),add=T,lwd=2,col="red")

hist(my.PRS.GW.STD[my.control.index],col=rgb(0,0,0,0.1),freq=F,breaks=7,xlim=c(-4,4),ylim=c(0,0.5),main="my.PRS.GW.STD",xlab="")
curve(dnorm(x,mean=mean(my.PRS.GW.STD[my.control.index]), sd=sqrt(var(my.PRS.GW.STD[my.control.index]))), add=T,lwd=2,col="black")
hist(my.PRS.GW.STD[my.case.index],col=rgb(1,0,0,0.1),breaks=7,freq=F,add=T)
curve(dnorm(x,mean=mean(my.PRS.GW.STD[my.case.index]), sd=sqrt(var(my.PRS.GW.STD[my.case.index]))),add=T,lwd=2,col="red")

hist(my.PRS.01.STD[my.control.index],col=rgb(0,0,0,0.1),freq=F,breaks=7,xlim=c(-4,4),ylim=c(0,0.5),main="my.PRS.01.STD",xlab="")
curve(dnorm(x,mean=mean(my.PRS.01.STD[my.control.index]), sd=sqrt(var(my.PRS.01.STD[my.control.index]))), add=T,lwd=2,col="black")
hist(my.PRS.01.STD[my.case.index],col=rgb(1,0,0,0.1),breaks=7,freq=F,add=T)
curve(dnorm(x,mean=mean(my.PRS.01.STD[my.case.index]), sd=sqrt(var(my.PRS.01.STD[my.case.index]))),add=T,lwd=2,col="red")

hist(my.PRS.01.null.STD[my.control.index],col=rgb(0,0,0,0.1),freq=F,breaks=7,xlim=c(-4,4),ylim=c(0,0.5),main="my.PRS.01.null.STD",xlab="")
curve(dnorm(x,mean=mean(my.PRS.01.null.STD[my.control.index]), sd=sqrt(var(my.PRS.01.null.STD[my.control.index]))), add=T,lwd=2,col="black")
hist(my.PRS.01.null.STD[my.case.index],col=rgb(1,0,0,0.1),breaks=7,freq=F,add=T)
curve(dnorm(x,mean=mean(my.PRS.01.null.STD[my.case.index]), sd=sqrt(var(my.PRS.01.null.STD[my.case.index]))),add=T,lwd=2,col="red")
```

## Performance of the PRSs, from the prediction perspective

\tiny
```{r,echo=FALSE,out.width = '90%',fig.align = "center"}
my.Y.cc=rep("0-Control",my.nsample); my.Y.cc[my.case.index]="1-Case"
my.ncase=sum(my.Y.cc=="1-Case") # total number of cases
my.ncontrol=sum(my.Y.cc=="0-Control") # total number of controls
a=seq(4,-2.5,by=-0.5)
P=rep(-9,length(a));TP=rep(-9,length(a));FP=rep(-9,length(a))
sensitivity=rep(-9,length(a)); specificity.1=rep(-9,length(a))

method=c("PRS.oracle","PRS.GW","PRS.01","PRS.01.null")
auc.method=rep(0,4)

PRS=my.PRS.oracle.STD
for (k in 1:length(a)) {
 PRS.threshold=a[k]
 P[k]=sum(PRS>PRS.threshold) 
 TP[k]=sum(my.Y.cc[PRS>PRS.threshold]=="1-Case") 
 FP[k]=sum(my.Y.cc[PRS>PRS.threshold]=="0-Control") 
 sensitivity[k]=TP[k]/my.ncase 
 specificity.1[k]=FP[k]/my.ncontrol 
}
plot(specificity.1,sensitivity,type="b",pch=20,col="gray",xlab="1 - Specificity = False Positives/ncontrols",ylab="Sensitivity = True Positives/ncases")
abline(0,1)
auc=0  
for (k in 2:length(a)) # Trapezoid's method height*(base1+base2)/2
  auc = auc + (specificity.1[k]-specificity.1[k-1])*(sensitivity[k]+sensitivity[k-1])/2
auc.method[1]=auc

PRS=my.PRS.GW.STD
for (k in 1:length(a)) {
 PRS.threshold=a[k]
 P[k]=sum(PRS>PRS.threshold) 
 TP[k]=sum(my.Y.cc[PRS>PRS.threshold]=="1-Case") 
 FP[k]=sum(my.Y.cc[PRS>PRS.threshold]=="0-Control") 
 sensitivity[k]=TP[k]/my.ncase 
 specificity.1[k]=FP[k]/my.ncontrol 
}
lines(specificity.1,sensitivity,type="b",pch=20,col="red")
auc=0  
for (k in 2:length(a)) # Trapezoid's method height*(base1+base2)/2
  auc = auc + (specificity.1[k]-specificity.1[k-1])*(sensitivity[k]+sensitivity[k-1])/2
auc.method[2]=auc

PRS=my.PRS.01.STD
for (k in 1:length(a)) {
 PRS.threshold=a[k]
 P[k]=sum(PRS>PRS.threshold) 
 TP[k]=sum(my.Y.cc[PRS>PRS.threshold]=="1-Case") 
 FP[k]=sum(my.Y.cc[PRS>PRS.threshold]=="0-Control") 
 sensitivity[k]=TP[k]/my.ncase 
 specificity.1[k]=FP[k]/my.ncontrol 
}
lines(specificity.1,sensitivity,type="b",pch=20,col="blue")
auc=0  
for (k in 2:length(a)) # Trapezoid's method height*(base1+base2)/2
  auc = auc + (specificity.1[k]-specificity.1[k-1])*(sensitivity[k]+sensitivity[k-1])/2
auc.method[3]=auc

PRS=my.PRS.01.null.STD
for (k in 1:length(a)) {
 PRS.threshold=a[k]
 P[k]=sum(PRS>PRS.threshold) 
 TP[k]=sum(my.Y.cc[PRS>PRS.threshold]=="1-Case") 
 FP[k]=sum(my.Y.cc[PRS>PRS.threshold]=="0-Control") 
 sensitivity[k]=TP[k]/my.ncase 
 specificity.1[k]=FP[k]/my.ncontrol 
}
lines(specificity.1,sensitivity,type="b",pch=20,col="green")
auc=0  
for (k in 2:length(a)) # Trapezoid's method height*(base1+base2)/2
  auc = auc + (specificity.1[k]-specificity.1[k-1])*(sensitivity[k]+sensitivity[k-1])/2
auc.method[4]=auc

legend(0.6, 0.37, paste(round(auc.method,3),method), lty=1, lwd=2, col = c("gray","red","blue","green"),cex=1.3)

#c("my.ncase=",my.ncase, "my.ncontrol=",my.ncontrol)
```

## \footnotesize Undertanding of the main diagonal line and AUC=50\% of a non-predictive PRS

\scriptsize
- Recall the scatter plot for the non-predictive PRS.0.01.null.STD
```{r,echo=FALSE,out.width = '30%',fig.align = "center"}
plot(my.PRS.01.null.STD,my.Y.STD,col=my.color.index)
abline(v=1,col="blue")
```

- Let $K<1$ be the population prevalence of the disease, so out of a total of $n$ samples, expect $n_{case}=n\cdot K$. 

- For each threshold $t$ used to call $P_t$ samples positives (cases), 

- Because the PRS used is not predictive, the expected true positives, $TP_t=P_t \cdot K$, and the expected 
$\mbox{sensitivity} =\frac{TP_t}{n_{case}} = \frac{P_t\cdot K}{n\cdot K} = \frac{P_t}{n}.$

- Similarly, the expected false positives, $FP_t=P_t \cdot (1-K)$, and the expected
$1-\mbox{specificity} =\frac{FP_t}{n_{control}} = \frac{P_t\cdot (1-K)}{n\cdot (1-K)} = \frac{P_t}{n}.$

- Thus, sensitivity ($x$) = 1-specificity ($y$) across the whole $\frac{P_t=0}{n}=0$ to $\frac{P_t=n}{n}=1$ range. That is, ROC of a non-predictive PRS is (expected) to be the main diagonal line ($x=y$), and AUC=50\%.

## \footnotesize Quiz: How can two PRSs with very different predictive performance be highly correlated?

\tiny  
```{r,echo=FALSE,out.width = '90%',fig.align = "center"}
par(mfrow=c(2,3))

plot(my.PRS.GW.STD,my.PRS.oracle.STD,col=my.color.index,xlim=c(-4,4),ylim=c(-4,4))
fit=lm(my.PRS.oracle.STD~my.PRS.GW.STD);abline(a=fit$coef[1],b=fit$coef[2],col="blue")
title(main=(paste(round(summary(fit)$coefficients[2,1],3)," (", signif(summary(fit)$coefficients[2,4],3),")",sep="")))

plot(my.PRS.01.STD,my.PRS.oracle.STD,col=my.color.index,xlim=c(-4,4),ylim=c(-4,4))
fit=lm(my.PRS.oracle.STD~my.PRS.01.STD);abline(a=fit$coef[1],b=fit$coef[2],col="blue")
title(main=(paste(round(summary(fit)$coefficients[2,1],3)," (", signif(summary(fit)$coefficients[2,4],3),")",sep="")))

plot(my.PRS.01.null.STD,my.PRS.oracle.STD,col=my.color.index,xlim=c(-4,4),ylim=c(-4,4)) 
fit=lm(my.PRS.oracle.STD~my.PRS.01.null.STD);abline(a=fit$coef[1],b=fit$coef[2],col="blue")
title(main=(paste(round(summary(fit)$coefficients[2,1],3)," (", signif(summary(fit)$coefficients[2,4],3),")",sep="")))

plot(my.PRS.01.STD,my.PRS.GW.STD,col=my.color.index,xlim=c(-4,4),ylim=c(-4,4))
fit=lm(my.PRS.GW.STD~my.PRS.01.STD); abline(a=fit$coef[1],b=fit$coef[2],col="blue")
title(main=(paste(round(summary(fit)$coefficients[2,1],3)," (", signif(summary(fit)$coefficients[2,4],3),")",sep="")))

plot(my.PRS.01.null.STD,my.PRS.GW.STD,col=my.color.index,xlim=c(-4,4),ylim=c(-4,4))
fit=lm(my.PRS.GW.STD~my.PRS.01.null.STD);abline(a=fit$coef[1],b=fit$coef[2],col="blue")
title(main=(paste(round(summary(fit)$coefficients[2,1],3)," (", signif(summary(fit)$coefficients[2,4],3),")",sep="")))

plot(my.PRS.01.null.STD,my.PRS.01.STD,col=my.color.index,xlim=c(-4,4),ylim=c(-4,4))
fit=lm(my.PRS.01.STD~my.PRS.01.null.STD);abline(a=fit$coef[1],b=fit$coef[2],col="blue")
title(main=(paste(round(summary(fit)$coefficients[2,1],3)," (", signif(summary(fit)$coefficients[2,4],3),")",sep="")))
```

## Recap the goal of this lecture: a **deeper** understanding of 
\small
\begin{itemize}  
\item the complexity of constructing a good PRS even under the simplest setting {\it without} LD or any heterogeneties.
\item the trouble introduced by false positives, due to multiple hypothesis testing and low power.
\item `the more is not always better' statement: PRS based on 6 gw-significant SNPs vs. 66 0.01-significant SNPS.
\item the various over-fitting or selection biases, $\hat \beta$ for a false positive or a true positive.
\end{itemize}

**What's next**  
\begin{itemize}
\item Effects of ex.nsample and ex.beta.true on AUC: easy to answer.
\item Answers to these Qs are less obvious: {\bf If we decrease ex.beta.true from 0.3 to 0.1 but increase ex.nsnp.true from 10 to 90},
\item[] \hspace{6pt} $h^2$ and SNP $h^2$? 
\item[] \hspace{6pt} AUC in general?
\item[] \hspace{6pt} AUC between PRS.gw and PRS.01?  
\end{itemize}
